Centering factor variables with interactions

### 2014-06-03 _knitr-posts/2014-05-30-centering-interactions.Rmd
Add renamed and updated centering post

---
layout: post
title: Centering, Interactions, and Model Averaging
subtitle: Why centering factor predictors can be important
published: false
---

@@

First of all, if you haven't already, go read the paper [Simple means to improve the interpretability of
regression coefficients](http://doi.org/10.1111/j.2041-210X.2010.00012.x) by Holger Schielzeth. This paper is a wealth of useful information. The code I've included below is just a simulated example to demonstrate an aspect of this paper.

@@

Now we make our predictions and plot the predictions:

@@

```{r, echo=FALSE}
# # Adjusting factor level references
#
# combine_coefs <- function(model, ref_coef, comp_coef) {
#   if(class(model)[1] == "lme")
#     x <- as.data.frame(summary(model)$tTable)
#   if(class(model)[1] == "glm" | class(model)[1] == "lm") {
#     x <- as.data.frame(summary(model)$coef)
#     names(x)[c(1, 2)] <- c("Value", "Std.Error")
#   }
#   x$coef_name <- row.names(x)
#   row.names(x) <- NULL
#   base_coef <- x[x$coef_name == ref_coef, "Value"]
#   base_se <- x[x$coef_name == ref_coef, "Std.Error"]
#   comparison_coef <- x[x$coef_name == comp_coef, "Value"]
#   comparison_se <- x[x$coef_name == comp_coef, "Std.Error"]
#   combined_coef <- base_coef + comparison_coef
#   combined_se <- sqrt(comparison_se^2 - base_se^2)
#   data.frame(coef = combined_coef, se = combined_se)
# }
#
# x2_coef <- combine_coefs(m2.2, ref_coef = "x1_cent", comp_coef = "x1_cent:x2_cent", unit_value = min(dat$x2_cent))
# x3_coef <- combine_coefs(m2.2, ref_coef = "x1_cent", comp_coef = "x1_cent:x3_cent", unit_value = min(dat$x3_cent))
#
# newdata$pred <- predict(m2.2, newdata = newdata)
# with(dat, plot(x1_cent + mean(dat$x1), y, col = f))
# plyr::d_ply(newdata, "x2_cent", function(x) {
#   with(x, lines(x1_cent + mean(dat$x1), pred, col = "lightgrey"))
#   })
#
# # very roughly:
# # see coef(m2.2)
# abline(a = 3, b = 0.6684 + 0.567*-0.333 + 0.988 * -0.3) # 1st level
# abline(a = 3, b = 0.6684 + 0.567*0.667 + 0.988 * -0.3) # 2nd level
# abline(a = 3, b = 0.6684 + 0.567*-0.333 + 0.988 * 0.7) # 3rd level

```

>>>>

---
layout: post
title: Centering factor variables with interactions
subtitle: Why centering factor predictors can be important
published: false
---

@@

First of all, if you haven't already, go read the paper [Simple means to improve the interpretability of
regression coefficients](http://doi.org/10.1111/j.2041-210X.2010.00012.x) by Holger Schielzeth. This paper is a wealth of useful information.

@@

Now we make our predictions and plot the predictions. We'll undo the centering of the `x1` variable here by adding the mean of the original data:

@@

# Calculating slopes and standard errors at interaction levels

Currently, our factor variable `f` has three levels. Our estimate `b1` represents the slope across the mean of the factors. However, we might want to present the slope with confidence intervals at each level of factor `f`.

Calculating the slope coefficients is fairly easy but the standard errors require a bit more math. The trick is to correctly combine the coefficient variances (or equivalently the squared standard errors) with the covariances between coefficients.

The way we combine the standard errors is the same way we combine the standard deviation of any normal distributions. For the case:

```
N1 ~ N(mu1, sigma1)
N2 ~ N(mu2, sigma2)
```

where `mu` represents the mean, `sigma` represents the standard deviation, and `N` represents a normal distribution.

We can calculate the combination of these two distributions multiplied by the constants `a` and `b` (which we'll call `N3`) as:

```
N3 = a * N1 + b * N2
N3 ~ N(mu3, sigma3)
mu3 = a * mu1 + b * mu2
sigma3 = sqrt(a^2 * sigma1^2 + b^2 * sigma2^2 + 2 * ab * cov(ab))
```

First, to illustrate combining the coefficient means, here's an example plotting the predictions by specifying the intercept and slopes. The first line in each of these `abline` function calls represents the intercepts and each factor level and the second line represents the slopes. Notice how we get these slopes by turning the two indicator columns to low or high values.

```{r cent4, fig.width=6, fig.height=5.5}
with(dat, plot(x1_cent, y, col = f))
b <- coef(m2.2)
se <- summary(m2.2)$coef[,2]
# The intercept and slope for the first level of the factor f:
abline(a = b[1] + b[3]*-0.333 + b[4] * -0.3, 
  b = b[2] + b[5]*-0.333 + b[6] * -0.3, col = 1)
# And the second level:
abline(a = b[1] + b[3]*0.667 + b[4] * -0.3, 
  b = b[2] + b[5]*0.667 + b[6] * -0.3, col = 2)
# And the third:
abline(a = b[1] + b[3]*-0.333 + b[4] * 0.7, 
  b = b[2] + b[5]*-0.333 + b[6] * 0.7, col = 3)
```

Before we calculate the standard errors on the slope at the three factor levels let's return to a more straightforward parameterization so we know what we're aiming for. I'll re-write the model here for clarity:

```{r, cent5, fig.width=6, fig.height=5.5}
m2 <- lm(y ~ x1 * f, data = dat)
b <- coef(m2)
se <- summary(m2)$coef[,2]

# Slopes:
with(dat, plot(x1, y, col = f))
abline(a = b[1], b = b[2], col = 1)
abline(a = b[1] + b[3], b = b[2] + b[5], col = 2)
abline(a = b[1] + b[4], b = b[2] + b[6], col = 3)

# SEs on those slopes:
# (as in Schielzeth 2010 Methods Ecol. Evol.)
se[2]
sqrt(se[5]^2 - se[2]^2)
sqrt(se[6]^2 - se[2]^2)
```

Now, let's derive those same standard errors for the centered version. We'll start be extracting the standard errors and covariances of the coefficients. Then we'll combine them:

```{r}
se <- summary(m2.2)$coef[,2]
x1x2_low <- min(dat$x2_cent)
x1x2_high <- max(dat$x2_cent)
x1x3_low <- min(dat$x3_cent)
x1x3_high <- max(dat$x3_cent)
vcov_x1_x1x2 <- vcov(m2.2)[5, 2]
vcov_x1_x1x3 <- vcov(m2.2)[6, 2]
vcov_x1x2_x1x3 <- vcov(m2.2)[6, 5]
se_x1 <- se[2]
se_x1x2 <- se[5]
se_x1x3 <- se[6]
```

Standard error on the slope at the first factor level:

```{r}
as.numeric(
sqrt(x1x2_low^2 * se_x1x2^2 + 
     x1x3_low^2 * se_x1x3^2 +
     se_x1^2 +
     2 * x1x2_low * vcov_x1_x1x2 +
     2 * x1x3_low * vcov_x1_x1x3 +
     2 * x1x2_low * x1x3_low * vcov_x1x2_x1x3)
)
```

This matches the version from the uncentered model:

```{r}
summary(m2)$coef[2,2] 
```

And the second level:

```{r}
as.numeric(
sqrt(x1x2_high^2 * se_x1x2^2 + 
     x1x3_low^2 * se_x1x3^2 +
     se_x1^2 +
     2 * x1x2_high * vcov_x1_x1x2 +
     2 * x1x3_low * vcov_x1_x1x3 +
     2 * x1x2_high * x1x3_low * vcov_x1x2_x1x3)
)
```

And the third level:

```{r}
as.numeric(
sqrt(x1x2_low^2 * se_x1x2^2 + 
     x1x3_high^2 * se_x1x3^2 +
     se_x1^2 +
     2 * x1x2_low * vcov_x1_x1x2 +
     2 * x1x3_high * vcov_x1_x1x3 +
     2 * x1x2_low * x1x3_high * vcov_x1x2_x1x3)
)
```

Again, comparing to the uncentered version, they match:

```{r}
se <- summary(m2)$coef[,2]
sqrt(se[5]^2 - se[2]^2)
sqrt(se[6]^2 - se[2]^2)
```
